---
title: "4, Code"
author: "Lynn Huang"
date: "August 17, 2020"
output: pdf_document
---
```{r setup, echo=FALSE}
rm(list=ls())
library(tidyverse)
library(dplyr)
setwd("C:/Users/lynn/OneDrive/Documents/ADULT/NCSU/ST558, Data Science for Statisticians")
```

#### 4.1 Manipulating Data: Logicals
Logical statements are comparisons between 2 quantities.
```{r}
"hi" == " hi"
"hi" == "hi"
4 == 1
4 != 1
```

Package tidyverse has package dplyr, useful for comparisons (esp. numeric).  
Due to loss of precision, first one will be FALSE even though it's TRUE!
```{r}
sqrt(3)^2 == 3
dplyr::near(sqrt(3)^2, 3)
```

Check type of object using is functions.
```{r}
is.numeric("Word")
is.numeric(10)
is.character("10")
is.na(c(1:2, NA, 3))
```
Can use Boolean vector to index (which elements to include/exclude).
Index using [], subset(), tidyverse/dplyr filter().
```{r}
# Tibble is like dataframe, just nicer printing properties
iris <- tbl_df(iris)
iris
# How to get only "setosa" irises?
iris$Species=="setosa"
iris[iris$Species=="setosa",]
subset(iris, Species=="setosa")
dplyr::filter(iris, Species=="setosa")    # NOTE: dplyr:: not necessary if only 1 filter() function
```
Beware implicit or explicit coercion, when R changes element type from less to more flexible: logical, integer, double, character.
```{r}
# Implicit coercion using c()
c("hi", 10)
c(TRUE, FALSE) + 0
c(TRUE, "hi")
mean(c(TRUE, FALSE, TRUE))

# Explicit coercion using as functions
as.numeric(c(TRUE, FALSE, TRUE))
as.character(c(1, 2, 3.5, TRUE))
```

Compound logic available with operators.
```{r}
# And using &. Or using |
set.seed(3)
(x <- runif(n=10, min=0, max=1))
(x < 0.25) | (x > 0.75)
# Double operators only check 1st comparison if given a vector
(x < 0.25) || (x > 0.75)

# Use logical operators to do multiple subsets on data
filter(iris, (Petal.Length>1.5) & (Petal.Width>0.3) & (Species=="setosa"))
iris[(iris$Petal.Length>1.5) & (iris$Petal.Width>0.3) & (iris$Species=="setosa"), ]

```

#### 4.2 Manipulating Data: Aside: R Packages
#### 4.3 Manipulating Data: dplyr Package
Lahman contains Major League Baseball data from 1871-2019 in 4 tables + others:  
* People = Player names, dates of birth, death, biographical info  
* Batting = Batting statistics  
* Pitching = Pitching statistics  
* Fielding = Fielding statistics  
Other tables about teams, post-season play, awards, Hall of Fame, etc. also included. Use `help(Lahman)` to see more details.  
```{r}
#install.packages("Lahman")
library(Lahman)
head(Batting, n=4)

# Tibble prints nicer
Batting <- tbl_df(Batting)
Batting

# Subset to get only PIT data, only 2000
filter(Batting, teamID=="PIT")
filter(Batting, teamID=="PIT" & yearID==2000)

# Re-order rows (default by ascending order)
arrange(Batting, teamID)
arrange(Batting, teamID, G)
arrange(Batting, teamID, desc(G))

# Subset columns: Select cols that match certain characteristic (contain X2B)
# $ operator returns simplified vector form, select() returns same type of object (tibble)
vec <- Batting$X2B
tib <- select(Batting, X2B)

# Piping/chanining can feed one function's output into another function's input
arrange(select(filter(Batting, teamID=="PIT"), playerID, G, X2B), desc(X2B))
Batting %>% filter(teamID=="PIT") %>% select(playerID, G, X2B) %>% arrange(desc(X2B))

# Select columns using multiple types of criteria
Batting %>% select(X2B:HR)
Batting %>% select(contains("X"))
Batting %>% select(starts_with("X"), ends_with("ID"), G)

# Rename variables
# NOTE: THis renaming isn't permanent, b/c we don't save output
Batting %>%
  select(starts_with("X"), ends_with("ID"), G) %>%
  rename("Doubles"=X2B, "Triples"=X3B)

# Re-order variables
# everything() grabs all other variables, so this puts playerID 1st, HR 2nd, then all other cols
Batting %>% select(playerID, HR, everything())
```

#### 4.4 Manipulating Data: Creating New Variables
From fivethirtyeight.com, fandango dataframe of film ratings (n=146 films, p=23 cols).
```{r}
#install.packages("fivethirtyeight")
library(fivethirtyeight)
fandango

# Add new column
fandango %>% mutate(avgRotten = (rottentomatoes + rottentomatoes_user)/2)
# Will be at the end (might have to scroll right in output), so select to view it up front
fandango %>% 
  mutate(avgRotten = (rottentomatoes + rottentomatoes_user)/2) %>%
  select(avgRotten)

# Transmute just grabs new column only (like mutate + select)
fandango %>% transmute(avgRotten = (rottentomatoes + rottentomatoes_user)/2)

# Summarize will apply basic functions like mean and sd to data
fandango %>% summarise(avgStars = mean(fandango_stars), sdStars = sd(fandango_stars))

# Can also summarize by group_by variable
fandango %>% group_by(year) %>% summarise(avgSTars = mean(fandango_stars), sdStars = sd(fandango_stars))
# NOTE: RUn this in console and you will see Groups: year[2] showing you've grouped by year into 2 groups

# Conditional Execution with If-Then-Else
# NOTE: Always GPP (good programming practice) to include base case if no conditions are met (else)
# How to create new variable for large setosa flowers?
# If can only take 1 comparison (see Warning), not meant for vectors
if ((iris$Petal.Length>1.5) & (iris$Petal.Width>0.3) & (iris$Species=="setosa")) {
  "Large Setosa"
}

# Use ifelse() function for a compound logical condition
help(ifelse)
ifelse((iris$Petal.Length>1.5) & (iris$Petal.Width>0.3) & (iris$Species=="setosa"), "L-S", "NotL-S")

# To save this new data label, use transmute() (mutate + selects new col to view) or mutate()
mutate(iris,
       Size=ifelse((Petal.Length>1.5) & (Petal.Width>0.3) & (Species=="setosa"), "L-S", "NotL-S"))

# Convert from wide to long data using gather() for machine learning
(tempsData <- read_delim(file="./Data/cityTemps.txt", delim=" "))
(newTempsData <- tempsData %>% gather(key=day, value=temp, 2:8))
# Convert form long to wide data using 
newTempsData %>% spread(key=day, value=temp)
# NOTE: Same data set as before, although columns now alphabetically ordered

# Split 1 col into multiple cols using separate()
(chicagoData <- read_csv(file="./Data/Chicago.csv"))
chicagoData %>% separate(date, c("Day", "Month", "Year"), sep="/")

```