---
title: "8, Code"
author: "Lynn Huang"
date: "September 7, 2020"
output: pdf_document
params:
  team: "Pittsburgh Steelers"
---
  
```{r setup, echo=FALSE}
rm(list=ls())
library(tidyverse)
library(knitr)
library(Lahman)
setwd("C:/Users/lynn/OneDrive/Documents/ADULT/NCSU/ST558, Data Science for Statisticians")
```

#### 8.1 Improving R Programs: Loops  
```{r FOR}
# FOR LOOP: For each index i in values 1:10, do something
for (i in 1:10){
  print(i)
}

# Values don't have to be numeric
for(index in c("cat", "hat", "worm"))
  print(index)

# Values don't have to be in any order
words <- c("first", "second", "third", "fourth", "fifth")
data <- runif(5)
# Want to print out sentence of format "The nth data point is #."
# Could require 5 lines of code, or 1 neat for-loop that iterates 5 times
paste0("The ", words[1], " data point is ", data[1], ".")
for (i in 1:5){
  print(paste0("The ", words[i], " data point is ", data[i], "."))
}

# Find summary() for each col in dataset
# Batting statistics for n=107429 observations on p=22 variables
#help(Batting)
newBatting <- Batting %>% as_tibble() %>% select(playerID, teamID, G, AB, R, H, X2B, X3B, HR)
newBatting
# Find summary() for a single column (col 3)
summary(newBatting[, 3])
# Find summary() for all cols using for-loop (skip char cols 1 and 2)
for (i in 3:dim(newBatting)[2]){
  print(summary(newBatting[, i]))
}
# Can also use seq_along() to get vector of ints referring to col nums
for (i in seq_along(newBatting)[-1:-2]){
  print(summary(newBatting[, i]))
}

# BREAK: Kicks you out of all loops
for (i in 1:5){
  if (i==3){
    break
  }
  print(paste0("The ", words[i], " data point is ", data[i], "."))
}
# Note that it works for all loops!
for (i in 1:5){
  for (j in -5:-1){
    if (j==-5) {break}
    print(paste0("j=", j))
  }
  print(paste0("i=", i))
}

# NEXT: Jump to next iteration of innermost loop

# Note that it only works on the innermost loop
for (i in 1:5){
  for (j in -5:-1){
    if (j==-5) {break}
    print(paste0("j=", j))
  }
  print(paste0("i=", i))
}
```

```{r WHILE}
# WHILE LOOP: While condition is TRUE, do something
i=1
while(i < 5){
  print(i)
  i <- i+1
}
```


#### 8.2 Improving R Programs: Vectorized Functions & Apply 
Loops are not the most efficient way when we want to work by row, by col, or by some other vectorized quantity! We'd prefer vectorized functions (later).  
Why are for-loops slow in R?  
1st, R must figure out what kind of object it is working with.  
2nd, R deploys the right function for your object type.  
For every iteration of the loop, it must repeat this as if the object were brand new, even if you are doing the exact same thing each iteration.  
```{r VectorizedFunc}
# Get some nice vectorized functions: colSds, colVars, colMedians
#install.packages("matrixStats")
library(matrixStats)

# Practice using built-in vectorized functions
colMeans(select(Batting, G:GIDP), na.rm=TRUE)

# Time small programs and get average run time over several (100s) runs
#install.packages("microbenchmark")
library(microbenchmark)

# Wrap code you want to microbenchmark
# Time in milliseconds, reported in 5-number summary for neval=100 runs
Bat <- Batting %>% select(G:GIDP)
microbenchmark(
  colMeans(Bat, na.rm=TRUE)
)
# Mean Time = 10.84047

# Compare vectorized function to for-loop
microbenchmark(
  for (i in 1:ncol(Bat)){
    mean(Bat[, i], na.rm=TRUE)
  }
)
# Mean Time = 35.58592 is much longer, so less efficient!

# How could we find median number of games (G), at bats (AB), and home runs (HR) for all players?
# Group observations by playerID, then summarize variables of interest (removing non-numeric col)
# Coerce to matrix, so you can apply colMedians() or summarise()
Batting %>% group_by(playerID) %>% summarise(totG=sum(G), totAB=sum(AB), totHR=sum(HR)) %>% 
  select(-playerID) %>% as.matrix() %>% colMedians(na.rm = TRUE)

# How could we find avg temp for different categories of wind speed?
# Daily air quality measurement sin New York from Day 1 to 153 (May to Sep in 1973) for n=153, p=6
data(airquality)
airquality <- as_tibble(airquality)
airquality

# Less Efficient: Loop thru each of n=153 obs of Wind and use ifelse to specify category
status <- vector()
for (i in seq_len(nrow(airquality))){
  if (airquality$Wind[i] >= 15){status[i] <- "HighWind"}
  else if (airquality$Wind[i] >= 10){status[i] <- "Windy"}
  else if (airquality$Wind[i] >= 6){status[i] <- "LightWind"}
  else if (airquality$Wind[i] >= 0){status[i] <- "Calm"}
  else {status[i] <- "Error"}
}
status
# Add status to airquality tibble
airquality$status <- status
# Can make calculations grouped by status, like mean temp
airquality %>% group_by(status) %>% mutate(avgTemp = mean(Temp))

# More Efficient: Use vectorized ifelse() to replace for-loop!
# Recall format ifelse(vector_condition, if_true_do_this, if_false_do_this)
status2 <- ifelse(airquality$Wind >= 15, 
                  "HighWind", 
                  ifelse(airquality$Wind >= 10, 
                         "Windy", 
                         ifelse(airquality$Wind >= 6, "LightWind", "Calm")))
# Note the perfect agreement in results (so same results, just different run times)
table(status, status2)

# Compare run times
loopTime<-microbenchmark(
  for (i in seq_len(nrow(airquality))){
  if (airquality$Wind[i] >= 15){status[i] <- "HighWind"}
  else if (airquality$Wind[i] >= 10){status[i] <- "Windy"}
  else if (airquality$Wind[i] >= 6){status[i] <- "LightWind"}
  else if (airquality$Wind[i] >= 0){status[i] <- "Calm"}
  else {status[i] <- "Error"}
  }
  , unit = "us")

vectorTime <- microbenchmark(
  ifelse(airquality$Wind >= 15, 
                  "HighWind", 
                  ifelse(airquality$Wind >= 10, 
                         "Windy", 
                         ifelse(airquality$Wind >= 6, "LightWind", "Calm")))
  , unit = "us")

loopTime
vectorTime
# This will be in nanoseconds.
mean(loopTime$time)/mean(vectorTime$time)
```

The family of apply() functions are not truly vectorized, but are also fast:  

* apply()  
* sapply()  
* lapply()  

There are others too, but we won't examine them in ST558.  
```{r APPLY}
# What is the apply() family of functions?
# Return vector/array/list of values obtained by applying FUN function to MARGIN of array/matrix X
help(apply)
# apply(X, MARGIN, FUN, ...) syntax. MARGIN=2 means by cols
apply(X=select(airquality, Ozone:Temp), MARGIN=2, FUN=summary, na.rm=TRUE)
# You can write your own functions too (unnamed function here b/c unnamed)
apply(X=select(airquality, Ozone:Temp), MARGIN=2,
			FUN = function(x) {
				temp <- c(mean(x, na.rm=TRUE), sd(x, na.rm=TRUE))
				names(temp) <- c("mean", "sd")
				temp
			}
)
# Note the mean values are the same in both apply() cases, as they should be!

# lapply(X, FUN, ...). Accepts list X, returns list of same length
help(lapply)
fit <- lm(Ozone ~ Wind, data = airquality)
fit <- list(fit$residuals, fit$effects, fit$fitted.values)
fit[[1]]
fit[[2]]
fit[[3]]

# What's the difference?
lapply(X = fit, FUN = mean)       # Returns list of 3 to correspond to input type
sapply(X = fit, FUN = mean)       # Returns vector of 3, which is simplified response type

# Not as efficient as vectorized functions, but better than for-loop!
air2 <- select(airquality, Ozone:Day)
microbenchmark(apply(X = air2, MARGIN = 2, FUN = mean, na.rm = TRUE))
microbenchmark(colMeans(air2, na.rm = TRUE))
```

#### 8.3 Improving R Programs: Parallel Computing
Often used in cross-validation or long computations where tasks need not be sequential.  
```{r Parallel}
# Use cross-validation to fit best classification model of survived or not on titanicData
# Contains information about n=1310 passengers, p=14 variables
# pclass = Categorical numeric for the passenger's fare class (1,2,3)  
# survived = Numeric factor where 0=died (lower level) and 1=survived (higher level)  
# name = Character  
# sex = Character (female/male)  
# age = Numeric (includes decimals)  
# sibsp = Numeric (number of siblings or spouses aboard)
# parch = Numeric (number of parents or children aboard)
# ticket = Numeric ticket number
# fare = Numeric fare paid
# cabin = Character cabin number
# embarked = Port of embarkation (C=Cherbourg, Q=Queenstown, S=Southampton)
# boat = Lifeboat number (if they survived, else NA)
# body = Body number (if they died and body was recovered, else NA)
# home.dest = Character (home destination)
titanicData <- read_csv("./Data/titanic.csv")
titanicData$survived <- as.factor(titanicData$survived)
titanicData$pclass <- as.factor(titanicData$pclass)
titanicData$sex <- as.factor(titanicData$sex)
titanicData

# MANUAL PARTITIONING: Kind of awful to do, so we'll use caret later
# Before dividing into k random folds for CV, set seed for replicability
set.seed(123)
n = dim(titanicData)[1]     # Number of passengers
k = 10                      # Number of folds (typically 10)
# Sample n indices randomly w/o replacement, then split them into folds
indices <- sample(n, replace = FALSE)
head(indices)
titanicData$fold <- vector(mode = "numeric", n)
# Assign foldnum for first 1 to n/k, then (n/k+1) to (2*n/k), then ... elements
for (foldnum in 1:k){
  i.first <- (n/k)*(foldnum-1) + 1
  i.last <- (n/k)*(foldnum)
  # Include next commented-out command if you want to see the 130 indices that go in each fold
  #print(indices[i.first:i.last])
  titanicData$fold[indices[i.first:i.last]] <- foldnum
}
titanicData <- filter(titanicData, !is.na(survived)) # Remove 1 row with survived=NA

# PARALLEL
# 1st, Write your own helper functions to split data, perform model fit, then get results
library(tree)
# Split data so kth fold will be test, the rest is train data
splitData <- function(data, foldnum) {
  train <- filter(data, fold != foldnum)
  test <- filter(data, fold == foldnum)
  # Can return multiple items as list
  list(train = train, test = test)
}
fitTitanicTree <- function(train, test){
  # Only numeric or pre-crash categorical variables used as predictors (no name, boat, body, home.dest)
  fit <- tree(survived ~ pclass + sex + age + ticket + fare + cabin + embarked, data = train)
  preds <- predict(fit, newdata = test, type = "class")
  misclass <- sum(abs(as.numeric(as.character(preds)) - as.numeric(as.character(test$survived))))
  # Can return multiple items as list
  return(list(fit = fit, preds = preds, misclass = misclass))
}
# Put previous 2 helper functions together into 1 big function to prep data, fit model, report results
getResult <- function(x, data){
  split <- splitData(data, foldnum = x)
  fitTitanicTree(split$train, split$test)  
}

# Check if helper functions work in single instance of non-parallel lapply()?
results <- lapply(1:k, FUN = getResult, data = titanicData)
results[[1]]$preds
results[[1]]$misclass
# Dig into results List and get misclassification rate for each of k=10 folds
# Function is '[' operator, which indexes in by "misclass" for each of the k=10 folds
# '[' is usually infix operator, but here is used as prefix operator (before args)
unlist(sapply(results, FUN = `[`, "misclass"))
# CV misclassificaton rate is sum of misclassifications across all k=10 folds (get back n obs total)
sum(unlist(sapply(results, FUN = `[`, "misclass")))/(n-1)
# 0.3147441

# 2nd, Ready to parallelize
library(parallel)
# Set up how many cores to use (leave at least 1 core for other computer tasks)
cores <- detectCores()
cores  
cluster <- makeCluster(cores - 1)
cluster
# On Windows, export packages, functions to cores
clusterExport(cluster, list("getResult", "splitData", "fitTitanicTree"))
clusterEvalQ(cluster, library(tidyverse))
clusterEvalQ(cluster, library(tree))

# 3rd, Use parLapply() instead of lapply()
resultsPar <- parLapply(cluster, X = 1:k, fun = getResult, data = titanicData)
sum(unlist(sapply(resultsPar, FUN = `[`, "misclass")))/(n-1)
# 0.3139801, similar to lapply() results! Prob only differs due to randomness

# Was it worth it? Check and see parallel time was less!
parTime <- microbenchmark(
  {resultsPar <- parLapply(cluster, X = 1:k, fun = getResult, data = titanicData)
  sum(unlist(sapply(resultsPar, FUN = `[`, "misclass")))/(n-1)},
  times = 10,
  unit = "s"
  )
# Might get some "NAs introduced by coercion" warning
straightTime <- microbenchmark(
  {results <- lapply(1:k, FUN = getResult, data = titanicData)
  sum(unlist(sapply(results, FUN = `[`, "misclass")))/(n-1)},
  times = 10,
  unit = "s"
  )
# Try this in console for nicest printing
print(parTime)
print(straightTime)
```

#### 8.4 Improving R Programs: Automating R Markdown
Go check out NFL.html file for sample output
```{r}
# Pull some NFL data
NFLData <- read_csv("data/scoresFull.csv")
NFLData
# Make same type of analysis file on Cleveland Browns or Pittsburgh Steelers (team param we want to set)
# NOTE: Such parameters can be specified in YAML header in the first place!
# "Knit with parameters" button will use specified params in YAML header
# Programatically automate with rmarkdown::render() and loop through set of params
rmarkdown::render("NFL.Rmd", output_file="Cleveland Browns.html", params=list(team="Cleveland Browns"))
data.frame(output_file="Pittsburgh Steelers.html", params=list(team="Pittsburgh Steelers"))

# How to make same type of analysis file on each unique team?
teamIDs <- unique(NFLData$awayTeam)
# Create filenames, then store in lists with just that team param stored
output_file <- paste0(teamIDs, ".html")
params = lapply(teamIDs, FUN = function(x){list(team = x)})
# Put into a data frame 
reports <- tibble(output_file, params)
reports

# Could use for-loop, but prefer apply() family to render() each output_file and param
library(rmarkdown)
# Index by [[]] to get at elements since tibble doesn't simplify (tibbles & dfs are just lists)
# Get rows (MARGIN=1) of reports, then take that row and apply render function to every row in reports
apply(reports, 
      MARGIN = 1,
      FUN = function(x){render(input = "NFL.Rmd", output_file = x[[1]], params = x[[2]])}
      )

# Tidyverse purr (replaced apply() family functions): pwalk (args are .l, .f, and ...)
#.l is a list of lists, .f is function, formula, or vector
pwalk(reports, render, input="NFL.Rmd")
```

